---
title: "BCB420 Assignment 1"
author: "Maryam Hasanzadehkiabi"
date: '2023-02-21'
output:
  html_document:
    toc: yes
    toc_depth: 2
  pdf_document:
    toc: yes
    toc_depth: '2'
subtitle: Dataset Selection and Initial Processing
bibliography: assignment_1.bib
csl: american-medical-association.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Dataset selection

## Insert and running required packages

[@xie2018r]
[@allaire2023rmarkdown]
[@xie2020r]
[@Zhu2021]
[@Müller2023]
[@Wickham2022]
[@Davis2007]
[@Morgan2022]
[@zhu2008geometadb]
[@durinck2009mapping]
[@durinck2005biomart]
[@drost2017biomartr]
[@loraine2015analysis]
[@robinson2010edger]
[@Wickham2022dev]
[@Dervieux2022]
[@Xie2023kni]
[@Xie2015kni]
[@Stodden2014]

```{r, message=FALSE, warning=FALSE}
if (!requireNamespace("rmarkdown", quietly = TRUE))
    +     install.packages("rmarkdown")
if (!requireNamespace("knitr", quietly = TRUE))
    +     install.packages("knitr")
if (!requireNamespace("pandoc", quietly = TRUE))
    +     install.packages("pandoc")
if (!requireNamespace("devtools", quietly = TRUE))
    +     install.packages("devtools")
if (!requireNamespace("BiocManager", quietly = TRUE))
    +     install.packages("BiocManager")
if (!requireNamespace("edgeR", quietly = TRUE))
    BiocManager::install("edgeR")
if (!requireNamespace("biomaRt", quietly = TRUE))
    BiocManager::install("biomaRt")
if (!requireNamespace("GEOmetadb", quietly = TRUE))
    BiocManager::install("GEOmetadb")
if (!requireNamespace("GEOquery", quietly = TRUE))
    +     install.packages("GEOquery")
if (!requireNamespace("DBI", quietly = TRUE))
    +     install.packages("DBI")
if (!requireNamespace("RSQLite", quietly = TRUE))
    +     install.packages("RSQLite")
if(!requireNamespace("kableExtra", quietly=TRUE))
    install.packages("kableExtra")
```

```{r message=FALSE, warning=FALSE}
library(devtools)
library(rmarkdown)
library(knitr)
library(pandoc)
library(BiocManager)
library(GEOmetadb)
library(GEOquery)
library(DBI)
library(RSQLite)
library(edgeR)
library(biomaRt)
library(kableExtra)
```

## Dataset Selection Criteria

1) RNASeq data 
2) Within the last 10 years
3) Limited to human
4) Number of samples more than 6
5) Related to Metabolic Syndrome

```{r message=FALSE, warning=FALSE}
sql <- paste("SELECT DISTINCT gse.title,gse.gse, gpl.title,",
             " gse.submission_date,",
             " gse.supplementary_file", 
             "FROM",
             "  gse JOIN gse_gpl ON gse_gpl.gse=gse.gse",
             "  JOIN gpl ON gse_gpl.gpl=gpl.gpl",
             "WHERE",
             "  gse.submission_date > '2013-01-01' AND", 
             "  gse.title LIKE '%Metabolic Syndrome%' AND", 
             "  gpl.organism LIKE '%Homo sapiens%' AND",
             "  gpl.technology LIKE '%high-throughput sequencing%' ",
             "  ORDER BY gse.submission_date DESC",sep=" ")
```

Selecting series with counts data

```{r, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
if(!file.exists('GEOmetadb.sqlite')) getSQLiteFile()
con <- dbConnect(SQLite(),'GEOmetadb.sqlite')
rs <- dbGetQuery(con,sql)
counts_files <- rs$supplementary_file[grep(rs$supplementary_file,
                              pattern = "count|cnt",ignore.case = TRUE)]

series_of_interest <- rs$gse[grep(rs$supplementary_file,
                              pattern = "count|cnt",ignore.case = TRUE)]

shortened_filenames <- unlist(lapply(counts_files,
             FUN = function(x){x <- unlist(strsplit(x,";")) ;
              x <- x[grep(x,pattern= "count|cnt",ignore.case = TRUE)];
               tail(unlist(strsplit(x,"/")),n=1)}))
shortened_filenames[1:10]
```

The output of the interested series resulted in only one series that holds the selection criteria. Therefore, in the following steps, GSE166474 [@li2021differentially] is being checked to find whether it contains sufficient samples to continue further analyses.

## Details of the Selected Series

Number of entities

```{r, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
gse <- getGEO("GSE166474",GSEMatrix=FALSE)
```

Data owner contact information

```{r, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
kable(data.frame(head(Meta(gse))), format = "pipe")
```

Name and number of the samples

```{r, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
names(GSMList(gse))
```
Series include 10 samples.

More information about the series

```{r, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
show(getGEO("GSE166474",GSEMatrix=TRUE))
```

Samples classification

```{r, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
data.frame(getGEO("GSE166474",GSEMatrix=TRUE)[[1]])[,c(8)]
```
There are 2 groups, 5 healthy/control individuals and 5 patients with metabolic syndrome. 

## Platform details

```{r, , echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
current_gpl <- names(GPLList(gse))[1]
current_gpl_info <- Meta(getGEO(current_gpl))
```

**Platform accession number:**
```{r, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
current_gpl_info$geo_accession
```
**Platform title:**
```{r, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
current_gpl_info$title
```
**Submission date:**
```{r, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
current_gpl_info$submission_date
```
**Last update date:**
```{r, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
current_gpl_info$last_update_date
```
**Organisms:**
```{r, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
current_gpl_info$organism
```
**Technology:**
```{r, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
current_gpl_info$technology
```
**Number of GEO datasets that use this technology:*
```{r, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
length(current_gpl_info$series_id)
```
**Number of GEO samples that use this technology:**
```{r, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
length(current_gpl_info$sample_id)
```

# 2. Data expression

Supplementary files

```{r}
supp_files = getGEOSuppFiles('GSE166474')
file_names = rownames(supp_files)
print(file_names)
```
There are two supplementary files provided for this dataset, one includes raw counts and the other one contains TPM normalized data [@li2021differentially]. Raw counts file is being used for further analysis. Researchers studied lncRNA, miRNA and mRNA differential expression of extracellular vesicles (EVs), and the effect of metabolic syndrome on their interactions [@li2021differentially]  

Displaying first 10 genes expression on 3 samples

```{r}
genes_expression = read.delim(file_names[1],header=TRUE,
          check.names = FALSE)
genes_expression[1:10, 1:4] %>%
    kableExtra::kbl(caption = "Genes Expression in 3 samples") %>%
    kableExtra::kable_classic(full_width = F, html_font = "Cambria")
```

In the following section, number of the genes and name of the variables included in the dataset are presented; since the first column includes gene IDs without column name, "ensembl_id" is assigned to the first column name.

```{r}
dim(genes_expression)
colnames(genes_expression)[1] <- "ensembl_id"
colnames(genes_expression)
```

Samples category

There are 2 groups included in the study, "M" indicates metabolic syndrome patients and "N" indicates healthy (normal) individuals. Each group contains 5 samples.

```{r}
dataSamples <- data.frame(lapply(colnames(genes_expression)[2:11], 
            FUN = function(x){unlist(strsplit(x,
                              split = "\\_"))[c(1)]}))
colnames(dataSamples) <- colnames(genes_expression)[2:11]
rownames(dataSamples) <- c("Status")
dataSamples <- data.frame(t(dataSamples))
```

# 3. Genes cleaning and filtering

## Gene name structure
Removing genes that do not follow ensembl nomenclature

```{r}
validGenes <- genes_expression[grep("ENSG", genes_expression$ensembl_id),]
```

## Duplication
Checking potential duplication. If the logical output returns any "TRUE" value, there does exist duplication. Otherwise the set contains unique genes.

```{r}
summary(duplicated(validGenes$ensembl_id))
```
Since there is no "TRUE" value in the output, dataset includes only unique genes.

## Outliers
Noise filtering is performed to keep only genes with at least 1 read per million in 5 samples (i.e. smallest group of replicates). However, since in this dataset, there are many gene expression with value equals to "0", further analyses, specifically log2(cpm) will cause (- infinity) value. 

```{r}
cpms = cpm(validGenes[,2:11])
rownames(cpms) <- validGenes$ensembl_id
keep = rowSums(cpms > 2) >=5
genes_filtered = validGenes[keep,]
filtered_results <- data.frame(genes_expression = nrow(genes_expression), validGenes = nrow(validGenes), genes_filtered = nrow(genes_filtered))
rownames(filtered_results)[1] <- "Number of Genes"
filtered_results %>%
  kableExtra::kbl(caption = "Genes count") %>%
  kableExtra::kable_classic(full_width = F, html_font = "Cambria")
```
In this dataset, there is 1 gene that does not follow the standard ensembl nomenclature, and there are 25,488 genes with low counts.

## Data distribution before normalization

Boxplot

```{r warning=FALSE}
data2plot <- log2(cpm(genes_filtered[,2:11]))
colnames(data2plot) <- lapply(colnames(genes_filtered)[2:11],
                              FUN=function(x) {unlist(strsplit(x,
                              split="\\_"))[c(1)]})
boxplot(data2plot, xlab = "Samples", ylab = "log2 CPM",
las = 2, cex = 0.5, cex.lab = 0.5,
cex.axis = 0.5, main = "Raw Data - Boxplot", col=rainbow(10))
abline(h = median(apply(data2plot, 2, median)),
col = "darkgreen", lwd = 1.2, lty = "dashed")
```

There are 3 outliers as negative infinity that are removed from plot. It seems that samples are very different.

Density plot

```{r}
counts_density <- apply(log2(cpm(genes_filtered[,2:11])),
2, density)
xlim <- 0; ylim <- 0
for (i in 1:length(counts_density)) {
xlim <- range(c(xlim, counts_density[[i]]$x));
ylim <- range(c(ylim, counts_density[[i]]$y))
}
cols <- rainbow(length(counts_density))
ltys <- rep(1, length(counts_density))
plot(counts_density[[1]], xlim=xlim, ylim=ylim, type="n",
ylab="Smoothing density of log2-CPM",
main="Raw Data Density Plot", cex.lab = 0.85)
for (i in 1:length(counts_density))
lines(counts_density[[i]], col=cols[i], lty=ltys[i])
legend("topright", colnames(genes_expression[2:11]),
col=cols, lty=ltys, cex=0.75,
border ="blue", text.col = "black",
merge = TRUE, bg = "gray90")
```
It seems that data follows a bimodal distribution pattern.

MA plot 

```{r}
plotMA(log2(genes_filtered[,c(2:6,7:11)]), ylab="M - ratio log expression",
main="MetS vs Control")
```

Scatter plot also does not show clustering around mean; there might be many outliers. 

# 4. Mapping

To perform mapping, ensembl_gene_id is selected

```{r}
ensembl <- useMart("ensembl")
ensembl = useDataset("hsapiens_gene_ensembl",mart=ensembl)
gene_mapped <- getBM(attributes = c("ensembl_gene_id","hgnc_symbol"),
filters = c("ensembl_gene_id"),
values = genes_filtered$ensembl_id,
mart = ensembl)
```


```{r}
length(which(rownames(cpms) %in%
gene_mapped$ensembl_gene_id))
```
Total number of genes with measurements
```{r}
nrow(cpms)
```
Number of genes that could not be mapped
```{r}
nrow(cpms) - length(which(rownames(cpms) %in%
gene_mapped$ensembl_gene_id)) 
```
Merging raw data and mapped genes; presenting data in one metobolic syndrome patient and one healthy person
```{r}
cpms_annot <- merge(gene_mapped,cpms,
by.x = 1, by.y = 0, all.y=TRUE)
kable(cpms_annot[1:6,c(1,2,3,8)],format = "html")
```
* Another method to check ensembl ID missing genes.
```{r}
length(ensembl_id_missing_gene <- cpms_annot$ensembl_gene_id[
which(is.na(cpms_annot$hgnc_symbol))])
```
Presenting the first 10 ensembl IDs that do not match with HUGO
```{r}
kable(cpms_annot[which(is.na(cpms_annot$hgnc_symbol)),1:3], format = "pipe")[3:12]
```

Old mapping

```{r}
old_mapping <- merge(genes_expression,data.frame(ensembl_id_missing_gene), by.x = 1, by.y = 1)
kable(old_mapping[1:10,], type="pipe")
```


# 5. Normalization

Checking potential duplication

```{r}
summary(duplicated(gene_mapped$ensembl_gene_id))
```

Removing duplication, and creating appropriate file to perform normalization analysis

```{r}
gene_mapped_unique <- unique(gene_mapped$ensembl_gene_id)
gene_mapped_unique <- data.frame(gene_mapped_unique)
rownames(gene_mapped_unique) <- gene_mapped_unique$gene_mapped_unique
gene_mapped_unique_counts = genes_expression[which(rownames(gene_mapped_unique)%in% genes_expression$ensembl_id),]
```
MA Plot

```{r}
plotMA(log2(gene_mapped_unique_counts[,c(2:6,7:11)]), ylab="M - ratio log expression",main="MetS vs Control")
```

Trimmed Mean of M-values (TMM)

```{r}
filtered_data_matrix <- as.matrix(gene_mapped_unique_counts[,2:11])
rownames(filtered_data_matrix) <- gene_mapped_unique_counts$ensembl_id
d = DGEList(counts=filtered_data_matrix, group = dataSamples$Status)
```

Normalization factors

```{r}
d = calcNormFactors(d)
kable(d$samples[,], format = "html") %>% row_spec(c(1,8,9,10), background = "yellow")
```
The normalization factor in the highlighted samples is less than 1. It means that small number of genes are expressed more in these samples; therefore, genes with less expression have higher relative value. 

```{r}
normalized_counts <- cpm(d)
```
```{r}
plotMDS(d, labels=lapply(rownames(dataSamples), FUN=function(x){unlist(strsplit(x,split = "\\_"))[c(1)]}), col = c("darkgreen","blue")[factor(dataSamples$Status)], main="MDS Plot")
```
Samples are not cluster with each other, there are unpaired samples that show different expression. It seems that N3, M1, M2, M3 and M6 samples could be less different. 

```{r}
model_design <- model.matrix(~dataSamples$Status)
```
```{r}
d <- estimateDisp(d, model_design)
```
```{r}
plotBCV(d,col.tagwise = "black",col.common = "red",col.trend = "blue", main="Biological Coeffient of Variation")
```
MDS plot is weierd and I could not explain the reason. 
Maybe that might be due to the study design [@li2021differentially], that to my understanding is a single cell analysis to detect variation of expression between cases and controls.

Normalized Data Boxplot

```{r warning=FALSE}
data2plotNorm <- log2(normalized_counts)
colnames(data2plotNorm) <- lapply(colnames(d)[1:10],
                              FUN=function(x) {unlist(strsplit(x,
                              split="\\_"))[c(1)]})
boxplot(data2plotNorm, xlab = "Samples", ylab = "log2 CPM",
las = 2, cex = 0.5, cex.lab = 0.5,
cex.axis = 0.5, main = "Normalized Data - Boxplot")
abline(h = median(apply(data2plotNorm, 2, median)),
col = "darkgreen", lwd = 1.2, lty = "dashed")
```

I used normalized data to check boxplot, but I could not explain what that weird boxplot means.

Normalized Data Density plot

```{r}
norm_counts_density <- apply(log2(normalized_counts[,]),
2, density)
xlim <- 0; ylim <- 0
for (i in 1:length(norm_counts_density)) {
xlim <- range(c(xlim, norm_counts_density[[i]]$x));
ylim <- range(c(ylim, norm_counts_density[[i]]$y))
}
cols <- rainbow(length(norm_counts_density))
ltys <- rep(1, length(norm_counts_density))
plot(norm_counts_density[[1]], xlim=xlim, ylim=ylim, type="n",
ylab="Smoothing density of log2-CPM",
main="Normalized Data Density Plot", cex.lab = 0.85)
for (i in 1:length(norm_counts_density))
lines(norm_counts_density[[i]], col=cols[i], lty=ltys[i])
legend("topright", colnames(genes_expression[2:11]),
col=cols, lty=ltys, cex=0.75,
border ="blue", text.col = "black",
merge = TRUE, bg = "gray90")
```
Density plot of normalized does not add any information and I could not explain why.

**What are the control and test conditions of the dataset?**
In this study, test condition includes 5 patients with metabolic syndrom, and controls include 5 healthy individuals.

**Why is the dataset of interest to you?**
Metabolic syndrome is defined as a set of metabolic disturbances including but not limited to, obesity, hypertension, insulin resistance that eventually results in cardiovascular disorders, liver disease, and type 2 diabetes. Many people worldwide are suffering from this condition [@li2021differentially].

**Were there expression values that were not unique for specific genes? How did you handle these?**
In this dataset, there is not any duplicate of genes. I used duplicated function to detect any duplication.

**Were there expression values that could not be mapped to current HUGO symbols?**
Yes. 24 genes are not mapped to the current HUGO symbols.

**How many outliers were removed?**
25,488 entries were removed with less than 1 read per million in the sample.

**How did you handle replicates?**
Except for those with low counts, replicates in the control and test conditions have been used in the data.

**What is the final coverage of your dataset?**
21.31%; that includes 6,903 genes, 10 samples (5 metabolic syndroms, and 5 healthy individuals)

All codes in this work are adapted from lectures instructed by Dr. Ruth Isserlin [@isserlin23]


## References













